\textbf{Dedukti: an implementation of set theory with pointed graphs} (titre provisoire)

### 2022-01-13 theoriezf/Article/Article.tex
Add files via upload

\begin{abstract}
We show how to implement a version of set theory in \dedukti ~in such a way that it respects normalization. To do so, we adapt in $\lambda \Pi$\textit{--calculus modulo theory} the guideline given in \textit{Deduction modulo} by G. Dowek and A. Miquel --using a theory of pointed graphs-- and we \textit{formally} write the proofs in \dedukti. To achieve this goal, we develop a language of formulas.
\end{abstract}

@@

\dedukti \ is a logical framework based upon the language \lpcm, which was created to overcome the lack of interoperability between proof assistants. As it is a logical framework, we can implement theories in \dedukti. Especially, set theory is interesting beacause it will enable traductions from and towards proof assistants using set theory, such as \textsc{Isabelle/ZF}.

A previous work by G. Dowek and A. Miquel \cite{zermodulo} has demonstrated the implemtation of set theory in \textit{Deduction modulo} thaks to a structure of \textit{pointed graphs}. Since the writing of \cite{zermodulo}, \lpcm \ has been created as an improvement of \textit{Deduction modulo}. We adapt the implementation to the \lpcm \ and then implement it in \dedukti.

@@

To implement set theory in \dedukti, the first idea is to translate the axioms of the theory into rewriting rules. However, as pointed out by M. Crabbé \cite{crabbé}, such implementation does not guarantee normalization. That is why G. Dowek and A. Miquel \cite{zermodulo} introduced a structure of pointed graphs to encode set theory. This set theory with pointed graphs -- called IZmod -- respects normalization.

@@

In the IZmod theory \cite[see][Section 3.1]{zermodulo}, sets are reprented by pointed graphs. A pointed graph is a direct graph whose one of its node is identified as the root.

@@

The operator $root$ gives the root of a pointed graph. $a/x$ replace the root of $a$ by $x$. $x~\eta_a~y$ means there is a edge in $a$ from $y$ to $x$. $rel(x,y,r)$ means that the nodes $x$ and $y$ are related by the relation $r$. We have the following rewriting rules:
$$ x \ \eta_{a/z} \ y \longrightarrow x \ \eta_a \ y $$
$$ \text{root}(a/x) \longrightarrow x $$
$$ (a/x)/y \longrightarrow a/y $$

We see in the examples that some different pointed graphs can represent the same set. To characterized this, G. Dowek and A. Miquel introduce bisimilarity between graphs, noted $\simeq$, and with rewriting rule

\begin{equation*}
\begin{split}
a \simeq b \lra~ &\ex~r~(rel(root(a),root(b),r) \\
&\conj \ \fa x \fa x'~\fa y~(x'~\eta_a~x~\conj~rel(x, y, r)~\imp~\ex~y'~(y'~\eta_b~y~\conj~rel(x', y', r))) \\
&\conj \ \fa y \fa y'~\fa x~(y'~\eta_b~y~\conj~rel(x, y, r)~\imp~\ex~x'~(x'~\eta_a~x~\conj~rel(x', y', r))))
\end{split}
\end{equation*}

@@

In set theory, there only exists one sort: the sets. In the IZmod theory, there exists two sorts: nodes and pointed graphs. Therefore, there are in IZmod two constructions related to the inclusion of sets. We can take the example of $\emptyset \in \{\emptyset\}$. We can represents this inclusion by two different ways.

@@

\node(a) at (3,2.75)[circle,draw,text centered,fill=gray] {h};
\node(k) at (3.5,2.75)[text centered]{$\in$};
\node(b) at (4,3)[circle,draw,text centered,fill=gray] {i};
\node(c) at (4,2.5)[circle,draw,text centered] {j};
\node(d) at (6,3)[circle,draw,text centered] {k};
\node(e) at (6,2.5)[circle,draw,text centered] {$\ell$};

\draw[->] (b) -- (c); 
\draw[->] (d) -- (e);

@@

Into a pointed graphs, inclusion is represented by the constructor $\eta$. Indeed, there exists an edge from $k$ to $\ell$, which means that the pointed graph with root $i$ is included in the pointed graph with root $h$.

Between two pointed graphs, inclusion is represented by a new constructor noted $\in$, with rewriting rule $a \in b \lra \ex~x~(x~\eta_b~root(b)~\conj~a \simeq (b/x))$. \\

The objective is to composed pointed graphs, for instance to join pointed graphs. The root of the new pointed graph is $o$. To guarantee that $o$ is not a node of one of the original pointed graphs, an injective function $i$ is introduced, such as its left inverse $i'$ and its image $I$. 
$$i'(i(x)) \lra x$$
$$ I(x) \lra \top $$
$$ I(o) \lra \bot $$

@@

Many definitions of the set theory intervines. The two classical set theories are Zermelo (Z) and Zermelo-Fraenkel (ZF). The set theory defined in \textit{Deduction modulo} by G. Dowek and A. Miquel \cite[see][Section 2]{zermodulo} --called IZst-- is stronger than Zermelo set theory but weaker than Zermelo-Fraenkel set thery. Indeed, the IZst theory contains Zermelo set theory, the Strong Extensionality axiom and the Transitive Closure axiom. \\

@@

The hypothesis of the Strong Extensionality axiom imitates the structure of the rewriting rule of $\eta$. This can be explained with the following reasoning: in the Extensionality axiom, the hypothesis states that two sets have the same elements, so the hypothesis of the strong extensionality axiom copy the structure of one of the equivalent to the inclusion in IZmod --that is to say the constructor $eta$. \\

@@

The Strong Extensionality axiom implies the Extensionality axiom\footnote{See \ref{extensionality}}.

@@

\node(a) at (2.25,3)[text width=3cm,text centered] {\textbf{Z} \\ Pairing \\ Union \\ Extensionality \\ Powerset \\ Comprehension \\ Infinity};
\node(b) at (6,3)[text width=10cm,text centered] {\textbf{IZst} \\ Transitive Closure \\ Strong Extensionality};
\node(c) at (10,3)[text width=3cm,text centered] {\textbf{ZF} \\ Replacement \\ Foundation};

@@

\subsection{From \textit{Deduction modulo} to \lpcm}

Since the publication of \cite{zermodulo}, \lpcm ~has been developped as an evolution of \textit{Deduction modulo}. More specifically, \lpcm ~allows to quantify on proposition. In that respect, the schemes of axioms have been replaced by axioms, as henceforth we do not need to write one axiom for each proposition. \\

In \dedukti, the symbol $mem$ and $rel$ \cite[see][Table 2]{zermodulo} are useless. Indeed, we can apply a node $x$ to class $P$ with $P~x$ and two nodes $x$ and $y$ to a relation $r$ with $r~x~y$.

The symbols $g_{x,y_1,...,y_n,P}$ and $g'_{x,x',y_1,...,y_n,P}$ had been created as specific constructors that allow to introduce in $mem$ a proposition $P$ by convert it into a class. These symbols are now not necessary in \dedukti since $mem$ is deleted. \\

The rewriting rule $$y = z \ \lra \ \fa \ p \ (mem(y, p) \ \imp \ mem(z, p))$$ is not required as it is a $mem$ is useless in \lpcm.

@@

All the other rewriting rules \cite[see][Table 3]{zermodulo}, except the ones involving $comp$, are easy to implement in \dedukti.

@@

The pointed graph $omega$ represents the set of Von Neumann ordinals \cite[see][Table 2]{zermodulo53}. Indeed, the informal presentation of the representation of sets by pointed graphs given in \ref{informal} corresponds to Von Neumann ordinals. The rewriting rule of $x~\eta_{omega}~x'$ states that in $omega$ the edges represents the order relation $<$ and that there is an edge from every natural number to the root of $omega$.

@@

We may have notice that the lemmas in which the $comp$ symbol is used are only valid for a subset of formulas \cite[see][Table 5]{zermodulo}. The formulas need to have all of its quantifiers of sort $El~graph$ and can only use the language $\in$, $\simeq$ and the classic logical connectives.

@@

There are 53 lemmas to proved \textit{formally} \cite[see][Tables a and 5]{zermodulo}.

The first lemma $x=x$ does not need to be implemented since it is already part of the standard library of \dedukti \ under the name $refl$ (which is polymorphic).

@@

This proof is written in \dedukti thanks to the following code:

@@

We implement it thanks to the $interpretation$ symbol. The valuation $update~\sigma~z~a)$ represents the assignement of variable $z \leftarrow a$.

@@

The proof of this $opaque~symbol$ is done by induction over formulas: each case is proved easily, using the lemmas that have already been checked by \dedukti. \\

@@

We follow the blueprint given by G. Dowek and A. Miquel \cite{zermodulo} : we use the strong extensionality axiom where $R(x,y)$ is $(x \simeq c~\conj~y \simeq d)~\disj~x \simeq y$. \\

@@

We have implemented in \dedukti ~a version of set theory -- IZst -- that respects normalization properties. IZst set theory correponds to Zermelo set theory, with the Strong Extensionality axiom and the Transitive Closure axiom. To do so, we have adapted the work by G. Dowek and A. Miquel \cite{zermodulo} from \textit{Deduction modulo} to \lpcm. We have \textit{formally} written all the proofs of the lemmas allowing us to implement set theory in \dedukti.

@@

>>>>

\begin{abstract}
We show how to implement a version of set theory in \dedukti ~in such a way that it respects normalization. To do so, we adapt in $\lambda \Pi$\textit{--calculus modulo theory} the guideline given in \textit{Deduction modulo theory} by G. Dowek and A. Miquel --using a theory of pointed graphs-- and we \textit{formally} write the proofs in \dedukti. To achieve this goal, we develop a language of formulas.
\end{abstract}

@@

Several prover are based on set theory, such as \textsc{Mizar}, \textsc{Atelier B} and \textsc{Isabelle/ZF}. The goal of this paper is to present an implementation of set theory in the logical framework \dedukti ~\cite{expressing} expressed in the \lpcm. We use here one of \dedukti's implementation, called \textsc{Lambdapi}. Our long term project is to use this implementation to enable interoperability between such provers based on set theory.

The first way to implement the set theory in \dedukti ~is to state each axiom in the framework. But such a method does not yield a cut-elimination theorem for the set theory. In particular, it does not enable to extract witnesses from a constructive existence proof. An alternative is to orient these axioms as rewriting rules. For instance, $$x \in \mathcal{P}(y) \Leftrightarrow x \subseteq y$$ is replaced with $$x \in \mathcal{P}(y) \lra x \subseteq y$$

However, as pointed out by M. Crabbé \cite{crabbé}, such a formulation of set theory does not have a cut-elimination property. An other solution is to define sets as \textit{pointed graphs}. With such a formulation, we can prove a cut-elimination theorem: for every proof in natural deduction that contains cuts there exists a cut-free proof. Such a theory expressed in \textit{Dedution modulo theory} is presented in \cite{zermodulo} with pencil and paper proofs. The goal of this paper is to present an implementation of this theory.

In 2007, the \lpcm ~had not been defined and this is why the original theory was defined in \textit{Dedution modulo theory}. The first issue was to adapt the expression in \textit{Dedution modulo theory} to the expression in \lpcm. This led to simplify the original formulation as classes of nodes of pointed graphs and relations on nodes can be naturaly expressed in \dedukti ~while that required a specific sort and specific comprehension axiom in \textit{Dedution modulo theory}. This enabled to reduce the size of the signature from 31 symbols to 26.

On the other hand, in the original formulation of the theory some lemmas were only valid for a specific subset of propositions. These lemmas were proved by induction over the structure of the propositions of this subset. Implementing these lemmas in \dedukti ~required to define a new inductive sort for the formulas of this subset and an intepretation of these formulas into the general type of  propositions, itself defined with rewriting rules. This development is the first, to our knowledge, to use such a reflexion principle defined with rewriting rules in \dedukti. The generality of this method still remains to be investigated.

@@

The set theory with pointed graphs developped by G. Dowek and A. Miquel is called IZmod (Intuitionistic Zermelo set theory in \textit{Deduction modulo}). We give here an informal presentation of the ideas developped in \cite{zermodulo}.

@@

In the IZmod theory, sets are reprented by pointed graphs. A pointed graph is a direct graph whose one of its node is identified as the root.

@@

The operator $root$ gives the root of a pointed graph. $a/x$ replace the root of $a$ by $x$. $x~\eta_a~y$ means there is a edge in $a$ from $y$ to $x$. The notation $rel(x,y,r)$ means that the nodes $x$ and $y$ are related by the relation $r$. We have the following rewriting rules:
$$ x \ \eta_{a/z} \ y \longrightarrow x \ \eta_a \ y $$
$$ root(a/x) \longrightarrow x $$
$$ (a/x)/y \longrightarrow a/y $$

We see in the examples that some different pointed graphs can represent the same set. To characterized this, a notion of  bisimilarity, noted $\simeq$, is introduced between graphs with rewriting rule

\begin{equation*}
\begin{split}
a \simeq b \lra~ &\ex~r~root(a)~root(b) \\
&\conj \ \fa x \fa x'~\fa y~(x'~\eta_a~x~\conj~r~x~y~\imp~\ex~y'~(y'~\eta_b~y~\conj~r~x'~y')) \\
&\conj \ \fa y \fa y'~\fa x~(y'~\eta_b~y~\conj~r~x~y~\imp~\ex~x'~(x'~\eta_a~x~\conj~r~x'~y')))
\end{split}
\end{equation*}

@@

In set theory, there only exists one sort: the sets. In the IZmod theory, there exists two sorts: nodes and pointed graphs. Therefore, there are in IZmod two constructions related to the inclusion of sets. We can take the example of $\emptyset \in \{\emptyset\}$. We can represents this inclusion by two different ways, one between pointed graphs and the other between nodes within a pointed graph.

@@

\node(d) at (6,3)[circle,draw,text centered] {k};
\node(e) at (6,2.5)[circle,draw,text centered,fill=gray] {$\ell$};
\node(k) at (6.5,2.75)[text centered]{$\in$};
\node(f) at (7,3)[circle,draw,text centered,fill=gray] {k};
\node(g) at (7,2.5)[circle,draw,text centered] {$\ell$};

\draw[->] (d) -- (e);
\draw[->] (f) -- (g);

@@

When in a graph we have an edge from $k$ to $\ell$, the pointed graph with the root $\ell$ represents a set which is an element of the set represented by the pointed graph with root $k$. But any pointed graph bisimilar with pointed graph with root $\ell$ also represents a set that is an element of the set represented with the pointed graph with root $k$. This leads to the definition of membership noted $\in$, with rewriting rule $a \in b \lra \ex~x~(x~\eta_b~root(b)~\conj~a \simeq (b/x))$. \\

The objective is to composed pointed graphs, for instance to join pointed graphs. The root of the new pointed graph is $o$. To guarantee that $o$ is not a node of one of the original pointed graphs, an injective function $i$ is introduced, with $i'$ its left inverse and $I$ the predicate of its image.

$$i'(i(x)) \lra x$$
$$I(x) \lra \top$$
$$I(o) \lra \bot$$

@@

These pointed graphs permit to prove the axioms of a theory IZst that lies between Zermelo (Z) and Zermelo-Fraenkel (ZF). This theory does not contain the Replacement scheme but it contains two additional axioms: the Strong Extensionality axiom and the Transitive Closure axiom. \\

@@

The hypothesis of the Strong Extensionality axiom imitates the structure of the rewriting rule of $\eta$. This can be explained with the following reasoning: in the Extensionality axiom, the hypothesis states that two sets have the same elements, so the hypothesis of the strong extensionality axiom copy the structure of one of the equivalent to the inclusion in IZmod --that is to say the constructor $\eta$. \\

@@

The Strong Extensionality axiom implies the Extensionality axiom\footnote{See Section \ref{extensionality}}.

@@

\node(a) at (2.25,3)[text width=3cm,text centered] {\underline{\textbf{Z}} \\ Pairing \\ Union \\ Extensionality \\ Powerset \\ Comprehension \\ Infinity};
\node(b) at (6,3)[text width=10cm,text centered] {\underline{\textbf{IZst}} \\ Transitive Closure \\ Strong Extensionality};
\node(c) at (10,3)[text width=3cm,text centered] {\underline{\textbf{ZF}} \\ Replacement \\ Foundation};

@@

\subsection{From \textit{Deduction modulo theory} to \lpcm}

Unlike the \textit{Dedution modulo theory}, the \lpcm ~allows quantification over propositions. \\

Thus the symbols $mem$ and $rel$ \cite[see][Table 2]{zermodulo} are not needed anymore. Indeed, a class of nodes can simply be expressed as a term $P$ of type $El~n \ra El~prop$, and the fact that a node $x$ is an element of this class can be expressed as $P~x$ and not as $mem(x,P)$. In a similar way, the fact that two nodes $x$ and $y$ are related by a relation $r$ is expressed with $r~x~y$.

The symbols $g_{x,y_1,...,y_n,P}$ and $g'_{x,x',y_1,...,y_n,P}$ had been created as specific constructors that allow to introduce in $mem$ a proposition $P$ by convert it into a class. These symbols are now not necessary in \dedukti ~since $mem$ is deleted.

@@

All the other rewriting rules \cite[see][Table 3]{zermodulo}, except the ones involving the symbol associated with the construction of sets by comprehension (noted $comp$), are easy to implement in \dedukti.

@@

The pointed graph \texttt{omega} represents the set of Von Neumann ordinals \cite[see][Table 2]{zermodulo53}. Indeed, the informal presentation of the representation of sets by pointed graphs given in \ref{informal} corresponds to Von Neumann ordinals. The rewriting rule of \texttt{eta omega x x'} states that in \texttt{omega} the edges represents the order relation $<$ and that there is an edge from every natural number to the root of \texttt{omega}.

@@

A key point in the original formulation of the theory is that the $comp$ symbol enabling to construct a set in comprehension is used are only valid for a subset of formulas \cite[see][Table 5]{zermodulo}. The formulas need to have all of its quantifiers of sort $El~graph$ and can only use the language $\in$, $\simeq$ and the classic logical connectives. Taking this restriction into account was the main challenge of this implementation.

@@

A first result on the original formulation of the theory was to prove that the axioms of IZst are theorems in IZmod. This required 53 lemmas that were \textit{informally} proved in \cite[see][Tables 4 and 5]{zermodulo} and that we proved \textit{formally} in this paper.

Some of these proofs just follow the informal ones. Some others required the use of the type of formula and its embedding into propositions.

The first lemma $x=x$ does not need to be implemented since it is already part of the standard library of \dedukti \ under the name \texttt{refl} (which is polymorphic).

@@

This proof is written in \dedukti ~thanks to the following code:

@@

We implement it thanks to the $interpretation$ symbol. The valuation $update~\sigma~z~a$ represents the assignement of variable $z \leftarrow a$.

@@

The proof of this \texttt{opaque symbol} is done by induction over formulas: each case is proved easily, using the lemmas that have already been checked by \dedukti. \\

@@

We follow the blueprint given by G. Dowek and A. Miquel \cite[see][Proposition 1]{zermodulo} : we use the strong extensionality axiom where $R(x,y)$ is $(x \simeq c~\conj~y \simeq d)~\disj~x \simeq y$. \\

@@

We have implemented in \dedukti ~a version of set theory -- IZst -- that respects normalization properties. IZst set theory correponds to Zermelo set theory, with the Strong Extensionality axiom and the Transitive Closure axiom. To do so, we have adapted the work by G. Dowek and A. Miquel \cite{zermodulo} from \textit{Deduction modulo theory} to \lpcm ~and have encoded set with a theory of pointed graphs -- IZmod. We have \textit{formally} written all the proofs of the lemmas allowing us to implement set theory in \dedukti.

@@

The implementation of IZmod theory represents a significant corpus of formal proofs in \textsc{Lambdapi}.

\newpage
\section*{Annex}
